
The main author and contributor to this thesis is Joel Sikström. Sections 1... were developed entirely by Joel, while Section 2 were developed in collaboration with Niclas Gärds and Casper Norrbin. Both Casper and Niclas have conducted a thesis on investigating usign allocators for use in ZGC in parallel with this thesis, but focusing on integrating an allocator and implementing/adapting a different allocator respectively. For the chapters develop in collaboration, Joel Sikström is the principal author of Section 2.1, 2.2 and 2.3, Casper the author of Section 2.4 and Niclas Section 2.5.

The main author and contributor to this thesis is Joel Sikström. All sections were developed entirely by Joel, except for Section~\ref{sec:background}, which were developed in collaboration with Niclas Gärds and Casper Norrbin. Both Niclas and Casper have conducted a thesis in the same subject in parallel with this one. Casper has conducted his thesis on the exact same subject as this thesis, but focusing on adapting a Buddy Allocator instead, which has resulted in a totally different set of adaptations and results. Niclas has conducted his work on integrating an allocator into ZGC, focusing on the challenges and opportunities of which is the natural future work of both Casper's and this thesis.

For the section developed in collaboration with Niclas and Casper, the contributions are as follows: Sections~\ref{sec:memory_management},~\ref{sec:fragmentation} and~\ref{sec:memory_allocation} are authored solely by Joel, Sections~\ref{sec:hotspot} and~\ref{sec:gc} are solely authored by Casper, and Section~\ref{sec:zgc} is authored solely by Niclas.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
